---
title: "Do male and female household heads spend differently on average on household welfare?"
author: "Hitoishi Chakma"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Per-capita food expenditure can serve as a proxy to understand consumption in the household: expenditure that directly relates to their wellbeing. 

Some G2P transfers, credit disbursements and pensions alike are directed towards female members of the household, under the assumption that this could increase agency of the female members and prompt the resources to directly contribute to the overall welfare of the household. This is in comparison to male members that may direct the resources towards expenses outside the household, without directly benefiting the household. 

Using data compiled from the 2009 National Income Dynamics Study (NIDS), we investigate whether there is a *measurable difference between per-capita food expenditure made by female and male household heads*. Answers to this question could have policy implications for development programs in their prioritizing/targeting of household heads.  

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
