---
title: "Do male and female household heads spend differently on average on household welfare?"
author: "Hitoishi Chakma"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Understanding per-capita food expenditure or consumption can serve as a proxy to understand household welfare. 

Some G2P transfers, credit disbursements and pensions alike are directed towards female members of the household, under the assumption that this could increase agency of the female members and prompt the resources to directly contribute to the overall welfare of the household. This is in comparison to male members that may direct the resources towards expenses outside the household, without directly benefiting the household. 

Using data compiled from the 2009 National Income Dynamics Study (NIDS), we investigate whether there is a *measurable difference between per-capita food expenditure made by female and male household heads*. Answers to this question could have policy implications for development programs in their prioritizing/targeting of household heads. 

```{r include=FALSE}
library(haven)       #Import Stata .dta 
library(labelled)
library(sjlabelled)
library(dplyr)       #Data manipulation
library(knitr)       #Tables for presentation. Kable function
library(data.table)  #Transpose
library(rstatix)     #Summary stats 
library(skimr)       #Get overview of data
library(janitor)
```

# Describing the Data

```{r echo=FALSE}
#read and store the data set
nids_dta <- read_dta("NIDS_Analysis_Final_2009.dta")

#change male column into factor with the value labels
nids_dta <- nids_dta %>% mutate(Gender = ifelse(male == 1, "Male", "Female"))
nids_dta$Gender <- as.factor(nids_dta$Gender)

#This is a fairly clean and labelled STATA file. Dictionary in Appendix 1
dictionary <- generate_dictionary(nids_dta)

nids_dta %>% 
  tabyl(Gender) %>% 
  kable()
  
```

The NIDS dataset contains responses from around 6886 individual household heads in South Africa of which 54.14% are male and 45.86% are female. 

```{r echo=FALSE}
Table_1_1 <- nids_dta %>% 
  tabyl(Gender, race) %>% 
  adorn_totals(where ="col") %>% 
  adorn_totals(where = "row") 
 
names(Table_1_1)[2:5] <- c("African", "Coloured", "Indian", "White")


# custom function to transpose 
transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df) 
}

Table_1_2 <- transpose_df(Table_1_1) 

names(Table_1_2) <- as.character(Table_1_2[1,]) # sorting the column names

Table_1_2 <- Table_1_2[-1,]

Table_1_2$Female <- as.numeric(Table_1_2$Female)
Table_1_2$Male <- as.numeric(Table_1_2$Male)
Table_1_2$Total <- as.numeric(Table_1_2$Total)

Table_1 <- Table_1_2

Table_1_Final <- kable(Table_1)

#Print the final table
Table_1_Final
```

Since this dataset has been compiled only on household heads, the mean age of the respondents is quite high. For males the mean age is around 48 years while for females it is 51 years.
Males in the data set seem to have completed around 1.5 years more of education than females on average. The average household size among the respondents is 4. However it should be noted that, female household heads tend to have larger households on average even when controlling for their race, age and education. Additionally, on average male household heads tend to earn more income per capita than female household heads.


```{r echo=FALSE}
#generate a few more descriptives
Table_2_1 <- nids_dta %>% 
                group_by(Gender) %>% 
                     summarise(
    "Mean Age" = round(mean(age, na.rm = TRUE), digits = 2),
    "Mean Years Education" = round(mean(education, na.rm = TRUE), digits = 2),
    "Mean HH Size" = round(mean(hhsize, na.rm = TRUE), digits = 2),
    "Mean PerCap Income (Rands)" = round(mean(hhpcinc, na.rm = TRUE), digits = 2),
    "Mean PerCap Food Expenditure (Rands)" = round(mean(pcfoodexp,na.rm = TRUE), digits = 2)
    ) 


Table_2_2 <- transpose(Table_2_1)
rownames(Table_2_2) <- colnames(Table_2_1)
colnames(Table_2_2) <- c("Female", "Male")
  
Table_2_3 <- Table_2_2[-1,]
Table_2_3$Female <- as.numeric(Table_2_3$Female)
Table_2_3$Male <- as.numeric(Table_2_3$Male)

Table_2_Final <- kable(Table_2_3)

Table_2_Final
```

The NIDS dataset contains responses from 6886 individual household heads in South
Africa of which 54.14% are male and 45.86% are female. Since this dataset has been compiled
only on household heads, the mean age of the respondents is quite high. For males the mean age
is around 48 years while for females it is 51 years.

Males in the data set have completed around 1.5 years more of education than females
on average. The average household size among the respondents is 4. Additionally, female household heads tend to have larger households on average even when controlling for their race, age and education. 

Important to note, on average male household heads tend to earn more income per capita than female household heads.^[this is a footnote]





## Appendix

### Appendix 1: Data Dictionary

```{r echo=FALSE}
dictionary
```

